---
title: "Sample summary"
output: 
  html_document:
      toc: true
      toc_float: true
      toc_depth: 4
      code_folding: NULL
      df_print: paged
editor_options: 
  chunk_output_type: editor
---


<!-- CSS style for left alignment and for tables to look ok: -->
<style>
body {
position: relative;
left: 42px;}
.main-container {max-width: none !important;}
table, td, th {
align: left;
font-size: 11px;
border: none;
padding-left: 5em;
padding-right: 5em;
min-width: 100%;
word-wrap: break-word;
max-width: 240px;
margin-left: auto;
margin-right: auto;
margin-top: 1em;
margin-bottom: 1em;
}
</style>

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE,comment = '',warning=F,error = F)
options(width = 1200)
```


# Dependencies

```{r, echo=F, include=FALSE}

# Depandencies
library(data.table)
library(ggplot2); theme_set(theme_bw())
library(patchwork)
library(stringr)
library(GenomicRanges)

```

# Parse data

```{r}

# Parse data
setwd('~/Data/CLINSEQ/frankenplot_test/')

files <- dir(pattern = '.plots.RDS')

# Tumor sample
t <- readRDS(files[1])


# Normal sample, if there is one


```

Gene shortcuts: AR [X], ATM [11], [PTEN](#chr10)

# Overview
```{r, echo=F, include=TRUE}

# Print tumor sample

p <- t

layout <-  "ABBBB
                CDDDD
                EFFFF
                GGGGG
                HHHHH
                IJJJJ
                IJJJJ"
        fig <- 
            p$gc_rawdepth+p$order_rawdepth+
            p$gc_log2+p$order_log2+
            p$depth_alleleratio+p$order_alleleratio+
            p$pos_log2+
            p$pos_alleleratio+
            p$nogrid+p$grid+
            plot_layout(design = layout,guides = 'collect')


print(fig)

```



# Genes


# Chromosomes

## Gene shortcuts

## 1 {#chr1} 

## 2 {#chr2}

## 3 {#chr3}

## 4 {#chr4}

## 5 {#chr5} 

## 6 {#chr6}

## 7 {#chr7} 

## 8 {#chr8} 

## 9 {#chr9} 

## 10 {#chr10}

## 11 {#chr11} 

## 12 {#chr12} 

## 13 {#chr13} 

## 14 {#chr14}

## 15 {#chr15}

## 16 {#chr16}

## 17 {#chr17} 

## 18 {#chr18} 

## 19 {#chr19} 

## 20 {#chr20} 

## 21 {#chr21} 

## 22 {#chr22} 

## X {#chrX}


























