---
title: "Introduction to Jumble"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to Jumble}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

**Jumble** is a tool for copy number analysis of short-read sequencing data. It supports both Targeted Sequencing (gene panels) and Whole Genome Sequencing (WGS).

## Installation

```r
# install.packages("devtools")
# devtools::install_github("your-repo/Jumble")
library(Jumble)
```

## Workflow Overview

1.  **Generate Count Files**: Generate binned read counts from BAM files.
2.  **Build Reference**: Create a reference panel from normal samples.
3.  **Run Analysis**: Normalize, segment, and call CNAs for tumor samples.

---

## Setup

```{r setup, warning=FALSE, message=FALSE}
library(Jumble)
library(data.table)

# Locate test data directory
testdata_dir <- system.file("testdata", package = "Jumble")
if (testdata_dir == "") testdata_dir <- "../inst/testdata" # Fallback for dev

print(testdata_dir)
```

## Workflow Steps

Jumble can be run entirely from R or using the included command-line wrapper scripts. The locations of the wrapper scripts can be found using:
```r
system.file("scripts", package = "Jumble")
```

### 1. Generate Count Files

The first step is to generate binned read counts from your BAM files. This step produces `.counts.RDS` files.

*   **For Reference Samples**: This step is necessary. You must generate count files for "normal" CNV-negative samples to build the reference panel in Step 2.
*   **For Test Samples**: This step is optional. You can run the analysis directly from BAM files (Step 3), in which case count files are generated automatically if needed. However, generating count files in advance may be useful for large cohorts.

This step requires a target BED file (for panels) or a bin size selection for WGS (10,000 recommended).

**R Function:**

```{r count_reads, eval=FALSE}
# Example for a list of BAM files
bam_files <- list.files("path/to/bams", pattern = ".bam$", full.names = TRUE)
target_bed <- "targets.bed"

# Process each BAM file
for (bam in bam_files) {
  generate_counts(
    bam_file = bam,
    target_bed_file = target_bed,
    output_dir = "counts_output"
  )
}
# This produces *.counts.RDS files
```

**Command Line Alternative:**

You can use the `jumble-count.R` wrapper script.

```bash
# Targeted Sequencing
Rscript jumble-count.R -t targets.bed -b sample.bam

# Whole Genome Sequencing (10kb bins)
Rscript jumble-count.R -w 10000 -b sample.bam
```

### 2. Build Reference

Once you have count files for your normal samples, you combine them into a reference file.

**R Function:**

```{r build_ref, warning=FALSE, message=FALSE}
# Locate example count files included in the package
count_files <- list.files(file.path(testdata_dir, "gene_panel/samples"), 
                          pattern = ".counts.RDS$", full.names = TRUE)

# In a real workflow, these would be your normal samples
print(basename(count_files))

# Build the reference
# We use 'annotation_source = "biomart"' to fetch gene data (requires internet)
# Note: For the sake of this vignette running offline, we skip the actual 
# biomaRt call by loading the pre-built reference, but the code below is how you run it.

# reference <- build_reference(
#  count_files = count_files,
#  annotation_source = "biomart",
#  genome = "hg19"
# )
```

**Command Line Alternative:**

Use `jumble-reference.R` to build a reference from a directory of count files.

```bash
Rscript jumble-reference.R -i reference_counts_dir/ -o reference.RDS
```

*Note: The `build_reference` function fetches gene annotations from Ensembl. To save time in this vignette, we will load the pre-built reference included in the test data for the next steps.*

```{r load_ref_real}
ref_file <- file.path(testdata_dir, "gene_panel/reference.RDS")
reference <- readRDS(ref_file)
```

### 3. Run Analysis

You can run the analysis starting directly from a BAM file (standard) or a pre-computed count file.

**R Function:**

#### Option A: Run from BAM (Standard)

```{r run_from_bam, eval=FALSE}
run_jumble(
  bam_file = "tumor_sample.bam",
  reference_file = reference,
  snp_vcf = "tumor_sample_germline_snps.vcf.gz",  # Optional: for GIS/LOH
  somatic_vcf = "tumor_sample_somatic.vcf.gz",  # Optional: for overlaying somatic mutations
  output_dir = "results"
)
```

#### Option B: Run from Counts (Example)

Here we run the analysis using the count file included in the test data.

```{r run_jumble, warning=FALSE, message=FALSE}
# Define input files
sample_file <- file.path(testdata_dir, "gene_panel/samples/test_sample_1.counts.RDS")

# Define optional VCF input
vcf_file <- file.path(testdata_dir, "gene_panel/samples/test_sample_1.vcf.gz")

# Create output directory
out_dir <- tempfile()
dir.create(out_dir)

# Run Jumble
results <- run_jumble(
  bam_file = sample_file,
  reference_file = ref_file,
  snp_vcf = vcf_file,
  output_dir = out_dir
)
```

#### Option C: Run with Somatic Annotations (Example)

You can optionally overlay somatic mutations onto the resulting plots by providing a somatic VCF file.

```{r run_jumble_somatic, eval=FALSE}
# Run Jumble with Somatic overlay
results_somatic <- run_jumble(
  bam_file = sample_file,
  reference_file = ref_file,
  snp_vcf = vcf_file,
  somatic_vcf = "path/to/somatic.vcf.gz",
  output_dir = out_dir
)
```

**Command Line Alternative:**

Use `jumble-run.R` for the complete analysis.

```bash
# Basic run
Rscript jumble-run.R -r reference.RDS -b tumor_sample.bam

# With SNP VCF (for GIS/LOH)
Rscript jumble-run.R -r reference.RDS -b tumor_sample.bam -v tumor_sample_germline_snps.vcf.gz

# With both SNP and Somatic VCFs
Rscript jumble-run.R -r reference.RDS -b tumor_sample.bam -v tumor_sample_germline_snps.vcf.gz -s tumor_sample_somatic.vcf.gz
```

### 4. Inspect Results

The `results` object is a list containing segments, bin-level data, and QC metrics.

```{r inspect_res}
# Segments (Copy Number Calls)
print(head(results$segments))

# QC Metrics
print(results$qc)
```

### 5. Genomic Instability Score (GIS)

If a SNP VCF is provided, Jumble calculates a GIS score to assess Homologous Recombination Deficiency (HRD). This is computed over a range of tumor fractions.

```{r inspect_gis}
if (!is.null(results$gis)) {
  print(head(results$gis))
}
```

### 6. Visualization

Jumble automatically generates plots in the output directory.

```{r list_plots}
list.files(out_dir, pattern = ".png")
```

---

# Citation

If you use Jumble in your research, please cite the following preprint:

> Mayrhofer, M. et al. "Sensitive detection of copy number alterations in samples with low circulating tumor DNA fraction." *MedRxiv* (2024). [doi:10.1101/2024.05.04.24306860](https://www.medrxiv.org/content/10.1101/2024.05.04.24306860v1)

# Conclusion

This workflow covers the basics of using Jumble for Copy Number Analysis. For WGS data, the process is identical, differing only in the `build_reference` step where you would typically use a fixed bin size (e.g., 10kb) rather than target regions.
